---
title: "SurvAlign Tutorial"
output: html_document
---

```{r preamble, echo=FALSE, warning=FALSE, message=FALSE}
devtools::load_all()
```

```{r load-lib}
library(survalign)

pid_colors <- c(
  "Democrat"="blue", 
  "Republican"="red", 
  "Independent/Other"="grey"
)
```

# Case Study: Measuring Group Alignment in the Cooperative Election Study (CES)

```{r load-ces-data}
data(ces)
```

## Exploring the Data

```{r check-ces-coverage}
ces |>
    plot_coverage(wave_col="year", ques_stem="immig")

ces |>
    plot_coverage(wave_col="year", ques_stem="spend")
```

```{r check-ces-support}
# Cross-sectional support 
ces |>
  filter(year == 2024) |>
  plot_support(weight_col="weight", group_col="pid3", ques_stem="immig") +
  ggtitle("Support for CES Immigration Items (2024)")
```

```{r check-ces-pairwise-support}
ces |>
  filter(pid3 == "Democrat", year == 2024) |>
  plot_pairwise_support(ques_stem="spend")
ces |>
  filter(pid3 == "Democrat", year == 2024) |>
  plot_pairwise_support(ques_stem="abort")
```

## Measuring Alignment

```{r write-ces-ques, echo=FALSE, results='asis'}
cat("\n#### CES Core Policy Preference Items\n")
ces_core_items <- colnames(ces)[grepl("(abort|immig|enviro|guns|military|trade)", colnames(ces))]
for (item in ces_core_items) {
  cat(paste0("\n- `", item, "`: ", attr(ces[[item]], "label")))
}
```

```{r measure-ces-pid-alignment, warning=FALSE, message=FALSE}
ces24_spend_align <- ces |>
  filter(year == 2024) |>
  measure_alignment(
    ques_stem="spend",
    group_col="pid3",
    id_col="id"
  )
ces24_spend_align |>
  plot_cumulative_support()
```

```{r plot-ces-pid-alignment, warning=FALSE, message=FALSE}
ces24_core_align <- ces |>
  filter(year == 2024) |>
  measure_alignment(
    ques_stem="(abort|immig|enviro|guns|military|trade)",
    group_col="pid3",
    id_col="id"
  )
ces24_core_align |>
  plot_group_alignment(show_dists_as="histogram")
```

```{r plot-ces-pid-alignment-curve}
ces24_core_align |>
  plot_alignment_curve(group_colors=pid_colors)
```

```{r summarise-ces-pid-alignment}
summary(ces24_core_align, format="long")
```

```{r plot-ces-pid-alignment-interactive}
ces24_core_align |>
    plot_alignment_curve(group_colors=pid_colors, interactive=TRUE)
```

## Alignment with Different Missing Value Assumptions

```{r ces-treat-na-as-unalign, warning=FALSE, message=FALSE}
ces24_core_align_na <- ces |>
  filter(year == 2024) |>
  measure_alignment(
    ques_stem="(abort|immig|enviro|guns|military|trade)",
    group_col="pid3",
    id_col="id",
    treat_na="unaligned"
  )

ces24_core_align_na |>
    plot_alignment_curve(group_colors = pid_colors)
```

## Alignment Over Time

varies considerably over time

```{r check-ces-support-over-time}
# Temporal support
ces |>
  plot_support(wave_col="year", weight_col="weight", 
               group_col="pid3", ques_stem="immig") +
  ggtitle("Support for CES Immigration Items")
```

```{r ces-pid-core-aligment-over-time, warning=FALSE, message=FALSE}
# Measure alignment over time
ces_pid_align <- ces |>
  measure_alignment_waves(
    ques_stem="(abort|immig|enviro|guns|military|trade)",
    group_col  = "pid3",
    wave_col   = "year",
    id_col     = "id",
    weight_col = "weight",
    verbose    = FALSE
  )
```

plot the alignment mean - nterpret it

```{r ces-alignment-mean-over-time}
ces_pid_align |>
  plot_group_stat_over_time(
    metric       = "alignment_mean",
    wave_col     = "year",
    group_col    = "pid3",
    group_label  = "Party ID",
    group_colors = pid_colors
  )
```
same story when looking at cumulative weak alignment

```{r ces-weak-alignment-over-time}
ces_pid_align |>
  plot_group_stat_over_time(
    metric       = "cumulative_weak_alignment",
    wave_col     = "year",
    group_col    = "pid3",
    group_label  = "Party ID",
    group_colors = pid_colors
  )
```

## Alignment on Different Issue Items

it may be sensitive to the choice of issue items

```{r write-ces-spend-ques, echo=FALSE, results='asis'}
cat("\n#### CES National Spending Items\n")
ces_spending_items <- colnames(ces)[grepl("spend", colnames(ces))]
for (item in ces_spending_items) {
  cat(paste0("\n- `", item, ": ", attr(ces[[item]], "label")))
}
```

```{r ces-pid-spending-aligment-over-time, warning=FALSE, message=FALSE}
# Plot support
ces |>
  plot_support(wave_col="year", weight_col="weight", 
               group_col="pid3", ques_stem="spend") +
  ggtitle("Support for CES Spending Items")

# Measure alignment
ces_spend_align <- ces |>
  measure_alignment_waves(
    ques_stem  = "spend",
    group_col  = "pid3",
    wave_col   = "year",
    id_col     = "id",
    weight_col = "weight",
    verbose    = FALSE
  )

# Plot alignment means
ces_spend_align |>
  plot_group_stat_over_time(
    metric       = "alignment_mean",
    wave_col     = "year",
    group_col    = "pid3",
    group_label  = "Party ID",
    group_colors = pid_colors
  )

# Plot cumulative weak alignment
ces_spend_align |>
  plot_group_stat_over_time(
    metric       = "cumulative_weak_alignment",
    wave_col     = "year",
    group_col    = "pid3",
    group_label  = "Party ID",
    group_colors = pid_colors
  )
```

## Alignment with Different Groups

```{r ces-educ-align, warning=FALSE, message=FALSE}
ces24_race_align <- ces |>
  filter(year == 2024) |>
  measure_alignment(
    ques_stem="(abort|immig|enviro|guns|military|trade)",
    group_col="race5",
    id_col="id"
  )

ces24_race_align |>
  plot_alignment_curve()
```

# Case Study: Measuring Group Alignment in the General Social Survey (GSS)

```{r load-gss}
data(gss)
```

```{r check-gss}
gss_cols <- colnames(gss)

gss_spending_items <- gss_cols[grepl("^nat.*_b", gss_cols)]
gss_social_items <- gss_cols[grepl(".*_b", gss_cols) & !grepl("^nat", gss_cols)]
```

```{r write-gss-ques, echo=FALSE, results='asis'}
cat("\n#### GSS National Spending Battery\n")
for (item in gss_spending_items) {
  cat(paste0("\n- `", item, "`: ", attr(gss[[item]], "label")))
}

cat("\n#### GSS Social Battery\n")
for (item in gss_social_items) {
  cat(paste0("\n- `", item, "`: ", attr(gss[[item]], "label")))
}
```

```{r plot-gss}
gss |>
    plot_coverage(wave_col="year", 
                  ques_cols=gss_spending_items)
gss |>
    plot_support(wave_col="year", 
                 weight_col="weight_sps",
                 ques_cols=gss_spending_items,
                 show_labels=FALSE)
```

Not all survey data comes with binarized questions. If so, the `binarize_responses()`
may come in handy.

```{r binarize-gss}
table(gss$natcity_b)
gss_b <- gss |>
    binarize_responses(ques_stem="^nat", threshold=0.5)
table(gss_b$natcity_b)
```


```{r gss-alignment-over-time, warning=FALSE, message=FALSE}
# Measure alignment
gss_spending_educ_align <- gss |>
  filter(year > 1972, !is.na(weight_sps)) |>
  measure_alignment_waves(
    ques_stem  = "nat.*_b",
    group_col  = "educ5",
    wave_col   = "year",
    id_col     = "id",
    weight_col = "weight_sps",
    verbose    = FALSE
  )

# Plot alignment means
gss_spending_educ_align |>
  plot_group_stat_over_time(
    metric       = "alignment_mean",
    wave_col     = "year",
    group_col    = "educ5",
    group_label  = "Education"
  )

gss_spending_educ_align |>
  plot_group_stat_over_time(
    metric       = "cumulative_weak_alignment",
    wave_col     = "year",
    group_col    = "educ5",
    group_label  = "Education"
  )
```

## GSS Alignment Over Time: Groups × Items × Measures {.tabset .tabset-pills}

```{r gss-multidim-setup, warning=FALSE, message=FALSE}
# Metrics configuration
ques_configs <- list(
  "Spending"      = list(ques_stem = "nat.*_b"),
  "Social Issues" = list(ques_cols = gss_social_items)
)

group_configs <- list(
  "Party ID"  = "pid3",
  "Education" = "educ5",
  "Race"      = "race3"
)

weight_configs <- list(
  "SPS Weight"  = "weight_sps",
  "NRPS Weight" = "weight_nrps"
)

# Measure alignment waves for each combination of items × groups × weights
gss_align_objects <- list()

for (ques_label in names(ques_configs)) {
  gss_align_objects[[ques_label]] <- list()
  for (group_label in names(group_configs)) {
    gss_align_objects[[ques_label]][[group_label]] <- list()
    for (weight_label in names(weight_configs)) {
      group_col  <- group_configs[[group_label]]
      weight_col <- weight_configs[[weight_label]]

      gss_dat <- gss |>
        filter(year > 1972, !is.na(.data[[weight_col]]))

      gss_align_objects[[ques_label]][[group_label]][[weight_label]] <-
        do.call(
          measure_alignment_waves,
          c(list(gss_dat,
                 group_col  = group_col,
                 wave_col   = "year",
                 id_col     = "id",
                 weight_col = weight_col,
                 verbose    = FALSE),
            ques_configs[[ques_label]])
        )
    }
  }
}
```

```{r gss-multidim-tabs, results='asis', echo=FALSE, warning=FALSE, message=FALSE}
# Estimate all metrics
gss_tab_metrics <- c("alignment_mean", "cumulative_weak_alignment")

all_lines <- character(0)
for (ques_label in names(gss_align_objects)) {
  all_lines <- c(all_lines,
    sprintf("### %s {.tabset .tabset-pills}", ques_label), "")
  for (group_label in names(gss_align_objects[[ques_label]])) {
    group_col <- group_configs[[group_label]]
    all_lines <- c(all_lines,
      sprintf("#### %s {.tabset .tabset-pills}", group_label), "")
    for (weight_label in names(gss_align_objects[[ques_label]][[group_label]])) {
      all_lines <- c(all_lines,
        sprintf("##### %s {.tabset .tabset-pills}", weight_label), "")
      for (metric in gss_tab_metrics) {
        metric_label <- stringr::str_to_title(stringr::str_replace_all(metric, "_", " "))
        all_lines <- c(all_lines,
          sprintf("###### %s", metric_label), "",
          "```{r, warning=FALSE, message=FALSE, echo=FALSE}",
          sprintf(
            'gss_align_objects[["%s"]][["%s"]][["%s"]] |> plot_group_stat_over_time(metric = "%s", wave_col = "year", group_col = "%s", group_label = "%s")',
            ques_label, group_label, weight_label, metric, group_col, group_label
          ),
          "```", "")
      }
    }
  }
}

cat(knitr::knit_child(text = all_lines, quiet = TRUE))
```
