---
title: "SurvAlign Tutorial"
output: html_document
---

```{r preamble, echo=FALSE}
devtools::load_all()
```

```{r load-lib}
library(survalign)

pid_colors <- c(
  "Democrat"="blue", 
  "Republican"="red", 
  "Independent/Other"="grey"
)
```

# Case Study: Measuring Group Alignment in the Cooperative Election Study (CES)

```{r load-ces-data}
data(ces)
```

## Exploring the Data

```{r check-ces-coverage}
ces |>
    plot_coverage(wave_col="year", ques_stem="immig")

ces |>
    plot_coverage(wave_col="year", ques_stem="spend")
```

```{r check-ces-support}
# Temporal support
ces |>
  plot_support(wave_col="year", weight_col="weight", 
               group_col="pid3", ques_stem="immig") +
  ggtitle("Support for CES Immigration Items")

# Cross-sectional support 
ces |>
  filter(year == 2020) |>
  plot_support(weight_col="weight", group_col="pid3", ques_stem="immig") +
  ggtitle("Support for CES Core Issue Items (2020)")
```

```{r check-ces-pairwise-support}
ces |>
  filter(pid3 == "Democrat", year == 2020) |>
  plot_pairwise_support(ques_stem="spend")
ces |>
  filter(pid3 == "Democrat", year == 2020) |>
  plot_pairwise_support(ques_stem="abort")
```

## Measuring Alignment

```{r write-ces-ques}
cat("\n#### CES Core Policy Preference Items\n")
ces_core_items <- colnames(ces)[grepl("(abort|immig|enviro|guns|military|trade)", colnames(ces))]
for (item in ces_core_items) {
  cat(paste0("\n- ", item, ": ", attr(ces[[item]], "label")))
}
```

```{r measure-ces-pid-alignment}
ces20_spend_align <- ces |>
  filter(year == 2020) |>
  measure_alignment(
    ques_stem="spend",
    group_col="pid3",
    id_col="id"
  )
ces20_spend_align |>
  plot_cumulative_support()
```

```{r plot-ces-pid-alignment}
ces24_core_align <- ces |>
  filter(year == 2024) |>
  measure_alignment(
    ques_stem="(abort|immig|enviro|guns|military|trade)",
    group_col="pid3",
    id_col="id"
  )
ces24_core_align |>
  plot_group_alignment(show_dists_as="histogram")
```

```{r plot-ces-pid-alignment-curve}
ces24_core_align |>
  plot_alignment_curve(group_colors=pid_colors)
```

```{r summarise-ces-pid-alignment}
summary(ces24_core_align, format="long")
```

```{r plot-ces-pid-alignment-interactive}
ces24_core_align |>
    plot_alignment_curve(group_colors=pid_colors, interactive=TRUE)
```

## Alignment with Different Missing Value Assumptions

```{r ces-treat-na-as-unalign}
ces24_core_align_na <- ces |>
  filter(year == 2024) |>
  measure_alignment(
    ques_stem="(abort|immig|enviro|guns|military|trade)",
    group_col="pid3",
    id_col="id",
    treat_na="unaligned"
  )

ces24_core_align_na |>
    plot_alignment_curve(group_colors = pid_colors)
```

## Alignment Over Time

```{r ces-pid-core-aligment-over-time}
# Measure alignment over time
ces_pid_align <- ces |>
  measure_alignment_waves(
    ques_stem="(abort|immig|enviro|guns|military|trade)",
    group_col  = "pid3",
    wave_col   = "year",
    id_col     = "id",
    weight_col = "weight",
    verbose    = FALSE
  )
```

```{r ces-alignment-mean-over-time}
ces_pid_align |>
  plot_group_stat_over_time(
    metric       = "alignment_mean",
    wave_col     = "year",
    group_col    = "pid3",
    group_label  = "Party ID",
    group_colors = pid_colors
  )
```

```{r ces-weak-alignment-over-time}
ces_pid_align |>
  plot_group_stat_over_time(
    metric       = "cumulative_weak_alignment",
    wave_col     = "year",
    group_col    = "pid3",
    group_label  = "Party ID",
    group_colors = pid_colors
  )
```

## Alignment on Different Issue Items

```{r write-ces-spend-ques}
cat("\n#### CES National Spending Items\n")
ces_spending_items <- colnames(ces)[grepl("spend", colnames(ces))]
for (item in ces_spending_items) {
  cat(paste0("\n- ", item, ": ", attr(ces[[item]], "label")))
}
```

```{r ces-pid-spending-aligment-over-time}
# Plot support
ces |>
  plot_support(wave_col="year", weight_col="weight", 
               group_col="pid3", ques_stem="spend") +
  ggtitle("Support for CES Spending Items")

# Measure alignment
ces_spend_align <- ces |>
  measure_alignment_waves(
    ques_stem  = "spend",
    group_col  = "pid3",
    wave_col   = "year",
    id_col     = "id",
    weight_col = "weight",
    verbose    = FALSE
  )

# Plot alignment means
ces_spend_align |>
  plot_group_stat_over_time(
    metric       = "alignment_mean",
    wave_col     = "year",
    group_col    = "pid3",
    group_label  = "Party ID",
    group_colors = pid_colors
  )

# Plot cumulative weak alignment
ces_spend_align |>
  plot_group_stat_over_time(
    metric       = "cumulative_weak_alignment",
    wave_col     = "year",
    group_col    = "pid3",
    group_label  = "Party ID",
    group_colors = pid_colors
  )
```

## Alignment with Different Groups

```{r ces-educ-align}
ces24_race_align <- ces |>
  filter(year == 2024) |>
  measure_alignment(
    ques_stem="(abort|immig|enviro|guns|military|trade)",
    group_col="race5",
    id_col="id"
  )

ces24_race_align |>
  plot_alignment_curve()
```

# Case Study: Measuring Group Alignment in the General Social Survey (GSS)

```{r load-gss}
data(gss)
```

```{r check-gss}
gss_cols <- colnames(gss)

gss_spending_items <- gss_cols[grepl("^nat.*_b", gss_cols)]
gss_social_items <- gss_cols[grepl(".*_b", gss_cols) & !grepl("^nat", gss_cols)]
```

```{r write-gss-ques}
cat("\n#### GSS National Spending Battery\n")
for (item in gss_spending_items) {
  cat(paste0("\n- ", item, ": ", attr(gss[[item]], "label")))
}

cat("\n#### GSS Social Battery\n")
for (item in gss_social_items) {
  cat(paste0("\n- ", item, ": ", attr(gss[[item]], "label")))
}
```

```{r plot-gss}
gss |>
    plot_coverage(wave_col="year", 
                  ques_cols=gss_spending_items)
gss |>
    plot_support(wave_col="year", 
                 weight_col="weight_sps",
                 ques_cols=gss_spending_items,
                 show_labels=FALSE)
```

Not all survey data comes with binarized questions. If so, the `binarize_responses()`
may come in handy.

```{r binarize-gss}
table(gss$natcity_b)
gss_b <- gss |>
    binarize_responses(ques_stem="^nat", threshold=0.5)
table(gss_b$natcity_b)
```


```{r gss-alignment-over-time}
# Measure alignment
gss_spending_educ_align <- gss |>
  filter(year > 1972, !is.na(weight_sps)) |>
  measure_alignment_waves(
    ques_stem  = "nat.*_b",
    group_col  = "educ5",
    wave_col   = "year",
    id_col     = "id",
    weight_col = "weight_sps",
    verbose    = FALSE
  )

# Plot alignment means
gss_spending_educ_align |>
  plot_group_stat_over_time(
    metric       = "alignment_mean",
    wave_col     = "year",
    group_col    = "educ5",
    group_label  = "Education"
  )

gss_spending_educ_align |>
  plot_group_stat_over_time(
    metric       = "cumulative_weak_alignment",
    wave_col     = "year",
    group_col    = "educ5",
    group_label  = "Education"
  )
```
